\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-27 coqimpl.tex
comments

This example might sound trivial and contrived. Let's look at the
example field "add". Now that we only expose the type of "add", we
cannot prove its commutativity in "comm_add" (because we don't have
access to its concrete definition) but we can override "add" with
subtraction function in "Family I2". This makes sense because we
shouldn't have (inherited) commutativity property for subtraction
anyway. This is contrary to how "a'" is declared "Final", the second "c"
can prove that it's equal to 1 but "a'" is not overridable in the
"Family I2".

@@

\textbf{Special Fields: ``Extensible'' Inductive Type and Defining Recursor.}
We don't actually support extensible inductive types from
\ref{langdesign-req2} in Coq (because that require non-trivial research
effort on theory side---i.e. {What is the semantics of extensible
inductive types?}---and engineering), but we simulate extensible
inductive type using overriding---when extending an inductive type $A$
with new constructors $c$, our plugin will generate a new inductive
definition $A'$ with this new $c$ on the fly and feed it to Coq.  But
still, we must be careful on the exposing data in the family type (the
compiled module type) for the following field.

@@

To construct a recursor as in \ref{langdesign-req3}, we create for each
constructor a handler and aggregate them inside a family. The whole
family "neg_handler" can be considered as a large matching case. 
Then we use
"FRecursor" to specify the motive and inductive type. Here "FRecursor"
will feed Coq
a module "v_33_34" for doing type-checking---what it does is basically
assert the existence of the recursor and apply to see if handlers can
pass the type checking by Coq.

Once recursor is created it will look like any other function fields (and compiled in the same manner), here we show "self_B.neg" to apply to "self_B.tt" for "B.example".  While in the implementation of our plugin, the concrete recursor for this "neg" will only be created during the compilation of the surrounding family into module---it will look exactly "term_for_type_checking" but uses concrete recursor. For example, at the appropriate place, our plugin will insert a definition into the compiled module: 
\begin{minted}{Coq}
  Definition neg := b_rec (fun _ => b) neg_handler.tt neg_handler.ff.
\end{minted}
where "b_rec" and "b" are the corresponding recursor and the (vanilla) inductive type in that compiled module.

Since the recursor is constructed knowing a concrete inductive type,
thus not inheritable, for all the children, we have to reconstruct and
override with a new recursor (therefore serving as exhaustiveness
checking for children families). 
This is affordable since each recursor handler doesn't need
to be rechecked---look at "B2.neg_handler", there is only one new
handler for "uu" and the handlers from the parent
"B.neg_handler.tt,B.neg_handler.ff" have already been type-checked and
inherited (since they only rely on ``the abstracted
interface'' module "b_3"). 
Thanks to family inheritance, their corresponding compiled module from the parents---for example, 
the module "ff_24"---doesn't require another type-check and is re-used during the type-checking and compilation of "B2.neg".

Of course, our plugin will ``inherit'' the
recursor "neg" in the surface 
but what the plugin does is doing a second exhaustiveness
checking in children family.\YZ{It'd be good to show *concretely* in Figure 4
(1) code that uses the recursor and how the using code is compiled, (2)
how B2 is compiled and that compiling B2 does not require recompiling
fields inherited from B.}\EDJreply{I added both. Please check.}

>>>>

This example might sound trivial and contrived. Let's look at the
example field "add".\YZ{This 'add' field is no longer there.} Now that we only expose the type of "add", we
cannot prove its commutativity in "comm_add" (because we don't have
access to its concrete definition) but we can override "add" with
subtraction function in "Family I2". This makes sense because we
shouldn't have (inherited) commutativity property for subtraction
anyway. This is contrary to how "a'" is declared "Final", the second "c"
can prove that it's equal to 1 but "a'" is not overridable in the
"Family I2".

@@

\textbf{Special Fields: ``Extensible'' Inductive Type and Defining Recursor.}
We don't actually support extensible inductive types from
\ref{langdesign-req2} in Coq (because that require non-trivial research
effort on theory side---i.e. {What is the semantics of extensible
inductive types?}---and engineering),\YZ{I do consider that your plugin supports extensible inductive types. It's just that they are implemented as a plugin, rather than in the Coq core.}
but we simulate extensible
inductive type using overriding---when extending an inductive type $A$
with new constructors $c$, our plugin will generate a new inductive
definition $A'$ with this new $c$ on the fly and feed it to Coq.  But
still, we must be careful on the exposing data in the family type (the
compiled module type) for the following field.

@@

To construct a recursor as in \ref{langdesign-req3}, we\YZ{A source of confusion in this section is that the reader often has to guess who performs the actions: is it the programmer or the Coq plugin?} create for each
constructor a handler and aggregate them into a family. The whole
family "neg_handler" can be considered as a large matching case. 
Then we use
"FRecursor" to specify the motive and inductive type. Here "FRecursor"
will feed Coq
a module "v_33_34" for doing type-checking---what it does is basically
assert the existence of the recursor and apply to see if handlers can
pass the type checking by Coq.

Once recursor is created it will look like any other function fields
(and compiled in the same manner). Here we show "self_B.neg" to apply to
"self_B.tt" for "B.example".
While in the implementation of our plugin, the concrete recursor for
this "neg" will only be created during the compilation of the enclosing
family into module---it will look exactly like "term_for_type_checking"
but uses concrete recursor. For example, at the appropriate place, our
plugin will insert a definition into the compiled module: 
\begin{minted}{Coq}
  Definition neg := b_rec (fun _ => b) neg_handler.tt neg_handler.ff.
\end{minted}
where "b_rec" and "b" are the corresponding recursor and the (vanilla) inductive type in that compiled module.

Since the recursor is constructed knowing a concrete inductive type,
thus not inheritable, for all the children, we have to reconstruct and
override with a new recursor (therefore serving as exhaustiveness
checking for children families).\YZ{Show this recursor, as you do above for the recursor "Definition neg := ...".}
This is affordable since each recursor handler doesn't need
to be rechecked---look at "B2.neg_handler", there is only one new
handler for "uu" and the handlers from the parent
"B.neg_handler.tt,B.neg_handler.ff" have already been type-checked and
inherited (since they only rely on ``the abstracted
interface'' module "b_3"). 
Thanks to family inheritance, their corresponding compiled module from the parents---for example, 
the module "ff_24"---doesn't require another type-check and is re-used during the type-checking and compilation of "B2.neg".

Of course, our plugin will ``inherit'' the
recursor "neg" in the surface 
but what the plugin does is doing a second exhaustiveness
checking in children family.

### 2022-06-27 lang-design.tex
comments

% We want to reason about fields in a family.
\textlabel{Req (1)}{langdesign-req1}:\textbf{ we want to be able to
reason about fields in a family}, just like in a module, where a field
can reason about its former fields. More generally, we want to allow a
later field to be type-dependent on the former fields. In the example of
\cref{fig:STLC-example}, "subst", "has_type", "subst_lemma" all require
this feature.\YZ{rephrase Req to Challenges}\EDJreply{But it is not really challenge because I didn't show their difficulty. It is more like requirements.}

@@

% we want to be able to reason pattern matching as well
However, there is still difference between data recursion and induction
reasoning, because the former one is \textit{proof-relevant}. This
difference leads to another issue: \textlabel{Req (4)}{langdesign-req4}:
\textbf{we need to reason about (the computation about) extensible
pattern matching}, just like "subst_lemma" in \cref{fig:STLC-example}.
When proving "subst_lemma", we have to know "subst" is invariant on
"tt", i.e., \mintinline{Coq}{(subst i x tt) = tt}. This kind of
information requires exposing \textit{computational rules} from the
recursors.\YZ{What goes wrong if subst is not invariant or you do not
know whether subst is invariant?}\EDJreply{Then it is unprovable because substitution lemma need to know this. You can just instantiate the whole lemma -- we want  
has_type (Γ; x ↦ A) tt T ->
has_type Γ u A ->
has_type Γ (subst i x tt) T. But this final goal requires to know the
computation of subst.}\YZreply{Seems like what's missing here is to
convince the reader why exposing the definition of subst could be a
problem for family polymorphism.}\EDJreply{Actually I don't see why I need to emphasize this "as a problem". This is just a requirement like any other requirements earlier. I don't know how to show its difficulty at this early stage.}

>>>>

% We want to reason about fields in a family.
\textlabel{Req (1)}{langdesign-req1}:\textbf{ we want to be able to
reason about fields in a family}, just like in a module, where a field
can reason about its former fields. More generally, we want to allow a
later field to be type-dependent on the former fields. In the example of
\cref{fig:STLC-example}, "subst", "has_type", "subst_lemma" all require
this feature.

@@

% we want to be able to reason pattern matching as well
However, there is still difference between data recursion and induction
reasoning, because the former one is \textit{proof-relevant}. This
difference leads to another issue: \textlabel{Req (4)}{langdesign-req4}:
\textbf{we need to reason about (the computation about) extensible
pattern matching}, just like "subst_lemma" in \cref{fig:STLC-example}.
When proving "subst_lemma", we have to know "subst" is invariant on
"tt", i.e., \mintinline{Coq}{(subst i x tt) = tt}. This kind of
information requires exposing \textit{computational rules} from the
recursors.\YZ{Seems like what's missing here is to
convince the reader why exposing the definition of subst could be a
problem for family polymorphism.}\EDJreply{Actually I don't see why I need to emphasize this "as a problem". This is just a requirement like any other requirements earlier. I don't know how to show its difficulty at this early stage.}\YZreply{There is apparently a tension between the need to expose definitional equality and the ability to override. Looking at the bigger picture, I think this section reads better if phrased as discussing challenges posed by adding family polymorphism to a proof assistant. I have some initial ideas that I need to discuss with you.}
